XSGRB - The X-shooter GRB afterglow sample

### 2017-01-15 document/XSGRB_sample.tex
Notes on bursts

\author{
	J.~Selsing\inst{1}, 
	D.~Malesani\inst{1}, 
	T.~Kr\"{u}hler\inst{1}, 
	J.~P.~U. Fynbo\inst{1}, 
	P.~Goldoni\inst{14}, 
	A.~de~Ugarte~Postigo\inst{11}, 
	J.~Japelj\inst{20},
	P.~D'Avanzo,
	Z.~Cano,
	S.~Covino\inst{10}, 
	V.~D'Elia\inst{7, 12}, 
	H.~Flores,
	O.~E.~Hartoog\inst{6},
	J.~Hjorth\inst{1}, 
	P.~Jakobsson\inst{5}, 
	A.~Levan,
	A.~Melandri,
	S.~Piranomonte \inst{7},
	R.~S\'anchez-Ram\'irez\inst{11},
	S.~Schulze\inst{17, 18}, 
	N.~R.~Tanvir\inst{19},
	C.~Th{\"o}ne,
	S.~D.~Vergani\inst{7, 8},
	P.~M.~Vreeswijk\inst{3}, 
	D.~J.~Watson\inst{1},
	K.~Wiersema\inst{19},
	D.~Xu\inst{1}
	L.~Christensen\inst{1},
	A.~De~Cia\inst{3}, 
	L.~Kaper\inst{6}, 
	L.~A.~Antonelli,
	F.~Fiore,
	A.~Gomboc,
	P.~Groot,
	F.~Hammer,
	C.~Ledoux\inst{2}, 
	E.~Maiorano,
	B.~Milvang-Jensen\inst{1}, 
	E.~Palazzi,
	E.~Pian,
	J.~Schaye,
	G.~Tagliaferri\inst{7},
	R.~A.~M.~J.~Wijers\inst{6}
}

@@

\institute{Dark Cosmology Centre, Niels Bohr Institute, University of Copenhagen, Juliane Maries Vej 30, 2100 K\o benhavn \O, Denmark
\and European Southern Observatory, Alonso de C\'{o}rdova 3107, Vitacura, Casilla 19001, Santiago 19, Chile
\and Department of Particle Physics and Astrophysics, Faculty of Physics, Weizmann Institute of Science, Rehovot 76100, Israel
\and Th\"uringer Landessternwarte Tautenburg, Sternwarte 5, 07778 Tautenburg, Germany
\and Centre for Astrophysics and Cosmology, Science Institute, University of Iceland, Dunhagi 5, IS-107 Reykjavik, Iceland
\and Astronomical Institute Anton Pannekoek, University of Amsterdam, Science Park 904, NL-1098 XH Amsterdam, the Netherlands
\and INAF-Osservatorio Astronomico di Roma, Via Frascati 33, I-00040 Monteporzio Catone, Italy
\and GEPI-Observatoire de Paris, CNRS UMR 8111, Univ. Paris-Diderot, 5 Place Jules Janssen - 92190 Meudon, France
\and American River College, Physics and Astronomy Dpt., 4700 College Oak Drive, Sacramento, CA 95841, USA
\and INAF, Osservatorio Astronomico di Brera, Via E. Bianchi 46, I-23807 Merate, Italy
\and Instituto de Astrof\'{\i}sica de Andaluc\'{\i}a (IAA-CSIC), Glorieta de la Astronom\'{\i}a s/n, 18008, Granada, Spain
\and ASI-Science Data Centre, Via Galileo Galilei, I-00044 Frascati, Italy
\and Institute of Experimental and Applied Physics, Czech Technical University in Prague, Horska 3a/22, 128 00 Prague 2, Czech Republic
\and APC, Astroparticules et Cosmologie, Universite Paris Diderot, CNRS/
	IN2P3, CEA/Irfu, Observatoire de Paris, Sorbonne Paris Cite, 10, Rue Alice Domon et L
	eonie Duquet, 75205 Paris Cedex 13, France
\and Max-Planck-Institut f\"{u}r extraterrestrische Physik, Giessenbachs
	tra\ss e, 85748 Garching, Germany
\and Universit\`a degli studi di Milano-Bicocca, Piazza della Scienza 3,
	20126, Milano, Italy
\and Pontificia Universidad Cat\'{o}lica de Chile, Departamento de Astro
	nom\'{\i}a y Astrof\'{\i}sica, Casilla 306, Santiago 22, Chile
\and Millennium Center for Supernova Science
\and Department of Physics and Astronomy, University ofLeicester, University Road, Leicester, LE1 7RH, UK
\and University of Ljubljana,Department of Physics,Faculty of Mathematics \& Physics,SI}

\abstract{
In this work we present the spectra of \textit{all} $\gamma$-ray bursts (GRBs)
afterglows that have \textit{ever} been observed at the VLT/X-shooter. This
totals 137 spectroscopic observation of 118 individual bursts, of which
redshifts have been secured for 91 \% covering a redshift range from 0.059 to
$\sim$ 8.5. Based on a set of unbiasing selection criteria, the follow-up effort
has been guided by the \textit{Swift}-telescope to produce a homogeneous sample
of afterglows for 86 spectra, representing the underlying \textit{Swift} GRB population.
We here provide a public release of all the spectra, including continuum
estimates and telluric corrections. We provide an assessment of the degree of
completeness with respect to the parent GRB population, in terms of the X-ray
properties of the bursts in the sample and find that the sample presented here
can not be significantly distinguished from the \textit{Swift} sample.
For all bursts (39) for which it is possible, we provide a measurement of the
optically derived hydrogen column density, increasing the total number of
published HI column densities by $\sim$ 50 \% and characterize the distribution
compared with literature values.
We constrain the fraction of dark burst to be 27\% - 42\% and confirm previous
results that higher optical darkness has increased X-ray excess
absorption.
%This dataset provides a unique possibility to study the ISM across cosmic time.
}

@@

For a brief moment, the energy output of a $\gamma$-ray burst (GRB) can rival
that of the stellar light integrated over the entirety of the observable
universe \citep{Kumar2015}. Bright across the entire electromagnetic spectrum,
the following afterglow makes possible the exploration of the intervening space,
otherwise dark. Since the discovery of the first $\gamma$-ray counterparts
nearly two decades ago\citep{Costa1997a, VanParadijs1997, Djorgovski1997,
	Frail1997}, significant advances in the understanding of the underlying
phenomenology and potential usage has been made.

@@

@@

%Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
%towards about 10$^5$ QSOs have 5 systems with
%$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
%\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
%systems in the majority of cases \citep{Jakobsson2006a, Fynbo2009}. Whereas DLAs
%towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
%atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
%\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
%star-forming galaxies over the full history of cosmic star formation from $z
%\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
%	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
%spectrum from X-rays to the sub-mm) we can hence characterize the properties of
%star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
%molecular contents, ISM temperatures, UV-flux densities, etc.. 
%This is, however,
%only possible as long as there are satellites in orbit that rapidly and
%accurately locate GRBs. The currently operating \textit{Swift} satellite,
%launched in 2004 and still fully-functioning, allows for very efficient
%follow-up observations of GRBs due to its unprecedented rate, speed and
%precision of localisations.

@@

In this section we describe how the final data products are produced and
subsequently post-processed. All post-processing scripts developed for this
dataset are made publicly available at
\url{https://github.com/jselsing/XSGRB_reduction_scripts}, along with
instructions of use.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
	\caption{Telluric corrected, normalized spectrum of GRB121024A at $z = 2.300$,
	meant to illustrate the typical data quality. The acquisition magnitude is r =
	20, meaning it is in the brighter end of the sample presented here, but not the
	brightest. The spectrum is rich in absorption, including molecular absorption
	of $H_2$ with the absorption due to \lya is visible at $\sim$ 4000 $\AA$.
	Additionally, three intervening systems are seen in this sightline  This
	spectrum has been analyzed in detail in \citet{Friis2015}. }
	\label{fig:spectrum}
\end{figure*}

@@

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 158
\todo{Get number} bursts has been triggered by \textit{Swift} in the period
since the commissioning of VLT/X-shooter in \todo{Insert date}. This sample
constitutes the \textit{statistical sample} on which we will try to address the
GRB population properties. Of this sample, 86 \todo{Get number} have been
spectroscopically followed up with X-shooter. In order to assess whether our
sample criteria and the subset of bursts followed up represent the underlying
parent population of GRB, we compare some intrinsic GRB properties of our sample
to the full sample of GRBs followed up by \textit{Swift}. Under the assumption
that \textit{Swift} randomly samples the underlying GRB population, if the
sample statistics of our subsample are similar to the \textit{Swift}-sample, we
can assume that the population properties should be conserved. There is some
evidence that the burst duration does not scale with redshift as expected
through time-dilation\citep{Kocevski2013, Littlejohns2013a}, pointing towards
that we could be missing an increasing fraction of the bursts at higher
redshifts, but \citet{Lien2016} argues that we should only be missing the bright
"short" GRBs", but to the degree that \textit{Swift} is complete, we can test if
the sample presented here is complete. We show the comparison between the burst
properties in Fig. \ref{fig:swift_complete}.

@@

A subset of all GRBs exhibit no detectable or very faint optical afterglow
\citep{Groot1998, Djorgovski2001, Fynbo2001}, which had been parametrized in
term of "darkness" \citep{Jakobsson2004, VanderHorst2009}. The X-ray properties
of this subset has previously been investigated \citep{DePasquale2003,
	Melandri2012}, finding a slightly higher X-ray luminosity compared to the
optically brights bursts combined with similar prompt characteristics pointing
towards a denser circumburst material. This indicates along with investigations
of host galaxy extinction \citep{Greiner2011, Kruhler2011, Hjorth2012}, that the
extinction of the optical afterglow is primarily driven by the presence of dust,
local to the burst \todo{Local dust reference}. Because the dust content of a
galaxy is correlated with the gas amount \citep{Bohlin1978, Guver2009}, we
expect the X-ray derived column densities for the bursts without an optical
afterglow to be higher than the ones without.

For all bursts with follow-up within 100 hours we can calculate the
"darkness"-parameter, $\beta_{OX}$ \citep{Milvang-Jensen2012}. This required the
simultaneous measurement of the X-ray flux density and the optical flux density
which is in practice never possible. We use the measured acquisition camera
magnitude reported in Tab. \ref{tab:sample_overview} to get the optical flux
density at the beginning of the spectroscopic integration. Because we know the
delay between the follow-up and the \textit{Swift} trigger, we can use the
measured XRT burst lightcurve \citep{Evans2007,
	Evans2009}\footnote{http://www.swift.ac.uk/xrt\_curves/trigger\_numer/flux.qdp}
to infer the corresponding X-ray flux density. This is done by either linearly
interpolating between temporally neighboring XRT measurements of by
extrapolating the last few X-ray data points to the spectroscopic delay. When
the afterglow is not detected in the acquisition camera, an upper limit of $> 24$
mags is used, which propagates into an upper limit on $\beta_{OX}$.

@@

Since \textit{all} GRBs should have an optical afterglow, the presence of an optical afterglow is more a statement about the sensitivity than optical properties of the bursts. \todo{Daniele, help.}

@@

Because the redshift completeness of our statistical sample is  54 percent
\todo{Get number}, making inference of the underlying redshift distribution of
GRBs is difficult. Because a large fraction of the bursts for which we were
unable to secure a redshift are missing due to terrestrial reasons, see Sect.
\ref{badbursts}, imposing additional unbiasing cuts on the sample can increase
the redshift completeness at the cost of sample size. As long as the cuts we
impose are not selective in terms of the burst properties, the sample properties
stay the same. The first cut we try is telescope availability such as bad
weather, visitor rejected the TOO, or unavailability of the instrument. Because
telescope availability is independent on the burst properties, by cutting away
these from our sample, the underlying redshift distribution should be conserved.
By trimming away the 30 bursts \todo{Get number} that can be rejected due to
telescope availability, the redshift completeness increases to 63 percent
\todo{Get number}. As is shown in Sect. \ref{timing}, the redshift completeness
decreases with increasing follow-up delay, we can additionally increase the
redshift completeness by imposing cuts on the follow-up delay. Our ability to
rapidly follow up a burst is also not dependent on the burst properties and
therefore we can additionally cut delays longer than 24 hours (all burst in the
sample are in principle observable within 24 hours).

@@

\subsection{GRB110709B (z=na)}

@@

\subsection{GRB111117A (z=1.3?)}

>>>>

\input{authorlist.tex}

@@

\abstract{
In this work we present the spectra of \textit{all} $\gamma$-ray bursts (GRBs)
afterglows that have \textit{ever} been observed at the VLT/X-shooter up to the end of 2016. This
totals 137 spectroscopic observation of 118 individual bursts, of which
redshifts have been secured for 91 \% covering a redshift range from 0.059 to
$\sim$ 8.5. Based on a set of unbiasing selection criteria, the follow-up effort
has been guided by the \textit{Swift}-telescope to produce a homogeneous sample
of afterglows for 86 spectra, representing the underlying \textit{Swift} GRB population.
We here provide a public release of all the spectra, including continuum
estimates and telluric corrections. We provide an assessment of the degree of
completeness with respect to the parent GRB population, in terms of the X-ray
properties of the bursts in the sample and find that the sample presented here
can not be significantly distinguished from the \textit{Swift} sample.
For all bursts (39) for which it is possible, we provide a measurement of the
optically derived hydrogen column density, increasing the total number of
published HI column densities by $\sim$ 50 \% and characterize the distribution
compared with literature values.
We constrain the fraction of dark burst to be 27\% - 42\% \todo{Preliminary} and tentatively confirm previous
results that larger optical darkness leads to increased X-ray absorption.
%This dataset provides a unique possibility to study the ISM across cosmic time.
}

@@

For a brief moment, the energy output of a $\gamma$-ray burst (GRB) can rival
that of the stellar light integrated over the entirety of the observable
universe \citep{Kumar2015}. Bright across the entire electromagnetic spectrum,
the following afterglow makes possible the exploration of the intervening space,
otherwise dark. Since the discovery of the first $\gamma$-ray counterparts
nearly two decades ago\citep{Costa1997a, VanParadijs1997, Djorgovski1997,
	Frail1997, Wijers1998}, significant advances in the understanding of the underlying
phenomenology and potential usage has been made.

@@

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006a, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities.

@@

@@

In this section we describe how the final data products are produced and
subsequently post-processed. $\sim$ 30 percent of the data released here have
been reduced in a manner similar to what is described in \citet{Kruhler2015}.
The two reduction schemes have been tested and are in good agreement. All
post-processing scripts developed for this dataset are made publicly available
at \url{https://github.com/jselsing/XSGRB_reduction_scripts}, along with
instructions of use.

@@

\begin{figure*}
	\centerline{\includegraphics[width=0.85\linewidth]{figures/GRB121024A.pdf}}
	\caption{Telluric corrected, normalized spectrum of GRB121024A at $z = 2.300$,
	meant to illustrate the typical data quality. The acquisition magnitude is r =
	20, meaning it is in the brighter end of the sample presented here, but not the
	brightest. The spectrum is rich in absorption, including molecular absorption
	of $H_2$. The absorption trough visible at $\sim$ 4000 $\AA$ is due to \lya in
	the host. We have overplot the most prominent lines seen in GRB afterglows from
	\citet{Christensen2011a}. The continuum estimate is shown in dashed red.
	Additionally, three intervening systems are seen in this sightline  This
	spectrum has been analyzed in detail in \citet{Friis2015}.}
	\label{fig:spectrum}
\end{figure*}

@@

Based on the sample criteria specified in Sect. \ref{samplecrit}, a total of 158
\todo{Get number} bursts has been triggered by \textit{Swift} in the period
since the commissioning of VLT/X-shooter in November, 2009. This sample
constitutes the \textit{statistical sample} on which we will try to address the
GRB population properties. Of this sample, 86 \todo{Get number} have been
spectroscopically followed up with X-shooter. In order to assess whether our
sample criteria and the subset of bursts followed up represent the underlying
parent population of GRB, we compare some intrinsic GRB properties of our sample
to the full sample of GRBs followed up by \textit{Swift}. Under the assumption
that \textit{Swift} randomly samples the underlying GRB population, if the
sample statistics of our subsample are similar to the \textit{Swift}-sample, we
can assume that the population properties should be conserved. There is some
evidence that the burst duration does not scale with redshift as expected
through time-dilation\citep{Kocevski2013, Littlejohns2013a}, pointing towards
that we could be missing an increasing fraction of the bursts at higher
redshifts, but \citet{Lien2016} argues that we should only be missing the bright
"short" GRBs", but to the degree that \textit{Swift} is complete, we can test if
the sample presented here is complete. We show the comparison between the burst
properties in Fig. \ref{fig:swift_complete}.

@@

A subset of all GRBs exhibit no detectable or very faint optical afterglow
\citep{Groot1998, Djorgovski2001, Fynbo2001}, which had been parametrized in
term of "darkness" \citep{Jakobsson2004, VanderHorst2009}. The X-ray properties
of this subset has previously been investigated \citep{DePasquale2003, Fynbo2009,
	Melandri2012}, finding a slightly higher X-ray luminosity compared to the
optically brights bursts combined with similar prompt characteristics pointing
towards a denser circumburst material. This indicates along with investigations
of host galaxy extinction \citep{Greiner2011, Kruhler2011, Hjorth2012}, that the
extinction of the optical afterglow is primarily driven by the presence of dust,
local to the burst \todo{Local dust reference}. Because the dust content of a
galaxy is correlated with the gas amount \citep{Bohlin1978, Guver2009}, we
expect the X-ray derived column densities for the bursts without an optical
afterglow to be higher than the ones without.

For all bursts with follow-up within 100 hours we can calculate the
"darkness"-parameter, $\beta_{OX}$ \citep{Jakobsson2004, Milvang-Jensen2012}. This required the
simultaneous measurement of the X-ray flux density and the optical flux density
which is in practice never possible. We use the measured acquisition camera
magnitude reported in Tab. \ref{tab:sample_overview} to get the optical flux
density at the beginning of the spectroscopic integration. Because we know the
delay between the follow-up and the \textit{Swift} trigger, we can use the
measured XRT burst lightcurve \citep{Evans2007,
	Evans2009}\footnote{http://www.swift.ac.uk/xrt\_curves/trigger\_numer/flux.qdp}
to infer the corresponding X-ray flux density. This is done by either linearly
interpolating between temporally neighboring XRT measurements of by
extrapolating the last few X-ray data points to the spectroscopic delay. When
the afterglow is not detected in the acquisition camera, an upper limit of $> 24$
mags is used, which propagates into an upper limit on $\beta_{OX}$.

@@

Since \textit{all} GRBs should have an optical afterglow, the presence of an optical afterglow is more a statement about the sensitivity than optical properties of the bursts. \todo{Hmm, how do I finish this?}

@@

Because the redshift completeness of our statistical sample is just 54 percent
\todo{Get number}, making inference of the underlying redshift distribution of
GRBs on this ample is difficult. Because a large fraction of the bursts for which we were
unable to secure a redshift are missing due to terrestrial reasons, see Sect.
\ref{badbursts}, imposing additional unbiasing cuts on the sample can increase
the redshift completeness at the cost of sample size. As long as the cuts we
impose are not selective in terms of the burst properties, the sample properties
stay the same. The first cut we try is telescope availability such as bad
weather, visitor rejected the TOO, or unavailability of the instrument. Because
telescope availability is independent on the burst properties, by cutting away
these from our sample, the underlying redshift distribution should be conserved.
By trimming away the 30 bursts \todo{Get number} that can be rejected due to
telescope availability, the redshift completeness increases to 63 percent
\todo{Get number}.

%As is shown in Sect. \ref{timing}, the redshift completeness
%decreases with increasing follow-up delay, we can additionally increase the
%redshift completeness by imposing cuts on the follow-up delay. Our ability to
%rapidly follow up a burst is also not dependent on the burst properties and
%therefore we can additionally cut delays longer than 24 hours (all burst in the
%sample are in principle observable within 24 hours).

@@

\subsection{GRB110709B (z=2.1092 (NEW!!!!))}
This is a late-time observation (> 1 year) and has previously been used in
\citet{Perley2016a}. In this reduction of the 7.2 ks spectroscopic integration,
the tentative detection of \oiii~reported in \citet{Perley2016a} is confirmed
along with low-significance detection of \ha~at the end of the spectral
coverage, both at a consistent redshift, $z=2.1092$, securing it as the redshift
of the GRB.

@@

\subsection{GRB111117A (z=2.211)}
